Data Analysis Project
=====================

Project preparation with ProjectTemplate [[1](#projecttemplate)].
``` {r knitr_options, echo=FALSE, warning=FALSE}
require(knitr)
opts_knit$set(root.dir="..")
opts_chunk$set(fig.path="../graphs/analysis-")
```

```{r load_project, echo=FALSE, warning=FALSE}
require(ProjectTemplate)
reload.project()
```

Introduction
------------

The project study the relationship between the highest degree earned by United
States residents and their family income in constant dollars.

Access to education and its funding is the subject of many discussions on social
mobility and redistribution of income. The study explores data from a long running
social survey to verify one of the main topic of these discussions: if family
income is related to education level.

Data
----

The study uses General Social Survey (GSS) data for the year 2012.

### General Social Survey (GSS)

The General Social Survey (GSS) has been monitoring change in American society
since 1972. Until 1994, it was conducted almost annually. Since 1994, the GSS has
been conducted in even numbered years.

The vast majority of GSS data is obtained in face-to-face interviews, computer-assisted
personal interviewing (since 2002), and by telephone.

The interviewees are individuals English and Spanish speaking persons 18 years of
age or older, living in the United States. They are selected from metropolitan and
rural areas. Multiple level of stratification for region, race, age, income and
sex was employed to garantee a random sample.

The target sample size is of 1500 observations every year until 1994. Since 1994 the GSS
has been administered to two samples in even-numbered years, each with a target
sample size of 1500.

For more information please refer to GSS FAQ [[2](#gssfaq)].

### Study characteristics

The data come from a survey and not from an experiment, so the study can be
characterized as **observational**: it can establish only correlation between
the variables examined and not causation.

However, GSS data are random samples taken from US residents, so the study's
findings could be **generalized** to the entire US residents population.

### Variables

The study uses two variables chosen from those collected by GSS survey:

* **Highest degree**: it is a ordinal categorical variable collected with the question
  "Did you ever get a high school diploma or a GED certificate?" -- its levels are
  *Less than high school*, *High school*, *Associate/Junior College*, *Bachelor's*,
  *Graduate*
* **Family income in constant US dollars**: it is a continuous numerical variable
  and it measure the inflation-adjusted family income

Exploratory data analysis
-------------------------

### Data preparation

Before exploring the data, we create a dataset with the two variables of interest
for the year 2012 and we give them a clearer label. We obtain a sample size of
1,974 observations.

```{r study_data, warning=FALSE}
# Filter Highest.Degree and Family.Income.Constant.USD for 2012
study_data = gss[gss$year == 2012, c(12,27)]
# Relabel the two variables
colnames(study_data) <- c("Highest.Degree","Family.Income.Constant.USD")
# Count the observations
nrow(study_data)
```

### Summary statistics

```{r summary_graphs, warning=FALSE, fig.width=12, fig.height=4, fig.cap='Summary statistics visualization'}
# Three graphs in a row
par(mfrow = c(1, 3))
# Barplot of Highest.Degree
par(mar=c(6,4,3,2))
barplot(table(study_data$Highest.Degree), las=2, main="Highest Degree")
# Histogram of Family.Income.Constant.USD
par(mar=c(6,1,3,2))
hist(study_data$Family.Income.Constant.USD, main="Family Income in constant USD", xlab="USD")
# Boxplot of Family.Income.Constant.USD by Highest.Degree
par(mar=c(6,2,3,4))
boxplot(study_data$Family.Income.Constant.USD ~ study_data$Highest.Degree, las= 2, main="Family Income in constant USD by Degree")
```

#### Highest degree

Highest degree is a categorical variable. We summarize it with a contingency table,
a frequency table and a barplot (see Figure 1).
```{r highest_degree_summary, warning=FALSE}
# Contingency table for Highest.Degree
table(study_data$Highest.Degree)
# Frequency table for Highest.Degree
prop.table(table(study_data$Highest.Degree))
```

We can see that high school as the highest degree has nearly 50% percent of observations.

#### Family income in constant USD

Family income in constant USD is a continuous numerical variable. We summarize
it with mean, range and quartiles, and with a histogram (see Figure 1).
```{r family_income_summary, warning=FALSE}
# Mean, range and quartiles for Family.Income.Constant.USD
summary(study_data$Family.Income.Constant.USD)
```

We can see that the distribution is right skewed and unimodal, with 50% of observations
in the 16,300-63,200 USD (constant dollar) range, and there is a maximum value
of 179,000 USD. There are some clear outliers in the upper quartiles of the distribution.

There are 216 observations with missing income values. Filtering them out brings
the number of observations to 1752. The sample size remains significant for the study.
```{r study_data_comp_observations, warning=FALSE}
# Filter NAs
study_data = study_data[complete.cases(study_data),]
# Count observations after the filter
nrow(study_data)
# Contingency table for Highest.Degree after the filter
table(study_data$Highest.Degree)
```

#### Relationship among family income and highest degree

Finally, we explore the relationship among family income and highest degree
(see Figure 1).

We can see that exists a positive association, but the wider interquantile range
in the college degree groups and the presence of outliers in the high school
degree and less than high school ones, means that such a relationship is
not strong and that family income could be associated with other variables.

Inference
---------

The study want to establish if there is a statistical significant difference
between the mean family income in constant dollars of United States resident
grouped by the highest degree earned.

In statistical inference terms, we test a null hypoteses ($H_{0}$) where the mean
family income in constant dollars is equal for all the highest degree groups,
and an alternative hypotheses ($H_{A}$) where at least one pair of means are different
from each other.

$H_{0}: \mu_{LHS} = \mu_{HS} = \mu_{JC} = \mu_{B} = \mu_{G}$

$H_{A}: the\ average\ income\ in\ constant\ dollar\ (\mu_{i})\ varies\ across\ some\ (or\ all)\ groups$

### ANOVA and its conditions

We could test the hypotheses doing a pairwise comparison of means across many groups.
But we could find a difference just by chance, even if there is no difference in
the population.

So we first use a method called **analysis of variance** (ANOVA) that uses a single
hypotheses test to check whether the mean across many groups are equal.
If we can reject the null hypotheses using ANOVA, then the results of pairwise
comparison are more significant.

ANOVA uses F test statistic, which represents a standardized ratio of variability
in the sample means relative to the variability within the group. The larger the
observed variability in the sample means relative to the within group observations,
the larger the F will be and the stronger the evidence against the null hypotheses.

ANOVA gives significant results if three conditions on the data are checked:

1. **indipendence**: *Are data indipendent?* -- GSS data consist in a random sample
   from less than 10% of the population and so they could be considered indipendent
2. **approximately normal**: *Have data a normal distribution?* -- normal probability
   plots for each group are shown in Figure 2, they visualize the difference among
   observations distribution and standard distribution; we can see that there is
   some deviation from normality in each group, especially in the upper quartiles

```{r qqplot_degree, warning=FALSE, fig.width=11, fig.height=2.5, fig.cap='Check data normality in family income grouped by degree'}
# Five graphs in a row
par(mfrow = c(1,5))
# Iterate on the groups and graph a QQ plot to test normality
degrees = c("Lt High School","High School","Junior College","Bachelor","Graduate")
for (i in 1:5) {
  qqnorm(study_data[study_data$Highest.Degree == degrees[i],]$Family.Income.Constant.USD, main=degrees[i])
  qqline(study_data[study_data$Highest.Degree == degrees[i],]$Family.Income.Constant.USD)
}
```

3. **constant variance**: *Is the variability across group about equal?* -- we can
   check the variability with the boxplot in Figure 1; we can see that the total
   range and the interquartile range of the groups are different, with the lowest
   variability in the Less than high school group and the highest variabilty in
   the Graduate group

The conditions on normality and constant variance are not fully respected.
We use ANOVA in our hypotheses test, but we report the uncertainty in the results.

### Computation

```{r anova, warning=FALSE}
# ANOVA for the mean income in constanst dollars grouped by degree
anova(lm(Family.Income.Constant.USD ~ Highest.Degree, data=study_data))
```

ANOVA reports a F test statistic of 121 and a p-value of approximately zero.
This mean that the probability of observing a F value of 121 or higher, if the null
hypotheses were true, is very low.

So we can reject the null hypotheses and we can say that the average income
in constant dollar varies across some (or all) groups of degree in a statistically
significant way.

We can do a pairwise comparison to find out which groups have different means.

For every possible pair of groups (10 pairs), we use a t test statistic to
confirm or reject the null hypotheses that the means of the two groups are equal 
($H_{0}: \mu_{Group1} = \mu_{Group2}$) or the alternative hypoteses that they are
different ($H_{A}: \mu_{Group1} \neq \mu_{Group2}$). The significance level
used is more stringent to avoid the increase of Type I error rate (rejecting
a true null hypotheses). This mean that to reject a null hypotheses the p-value
has to be much lower.

```{r pairwise, warning=FALSE}
# Pairwise t test for the mean income in constanst dollars grouped by degree
# With Bonferroni correction
pairwise.t.test(study_data$Family.Income.Constant.USD, study_data$Highest.Degree,  p.adj="bonferroni")
```

We can see tha for nine group pairs the p-value is lower than the significance
level and so the null hypothese is rejected: the difference of their means is
statistically significant.

The only pair for which the null hypotheses is not rejected is High School --
Junior College. Stated otherwise, the means of income in constant dollars for
the High School degree and the Junior College Degree are equals, and every
variation is due to chance.

Conclusion
----------

The study wanted to answer to the question if a higher family income is related
to an higher education level.

Using statistical inference methods, we find that this is the case.
The mean family income is significantly different between those who have
lt high school, high school, junior college, bachelor, graduate.
The only two groups that don't have significant difference are high school and
junior college.

The data used are from GSS a long running social research and it could be
generalized to all United States residents.

There are some conditions not meet in the normality and constant variance of the
data and so the results could not be considered definitive. Further analysis
with more sofisticated statistical method could be made to obviate to the problem
in the data.

Outliers in groups. Probably other variables could explain income.

Moreover it could be interesting doing the study for various year in the GSS survery
to find out if the difference in the mean evolved during the years.


#### Question
Write a brief summary of your findings without repeating your statements from earlier. Include a discussion of what you have learned about your research question and the data you collected, and include ideas for possible future research.

#### Evaluation
Is there a brief summary of findings that does not repeat previous statements?

Is a discussion of what was learned about the research question provided?

Are ideas for possible future research and/or discussion of additional synthesis or possible shortcomings of study provided?

References
----------

### Data reference

General Social Survey Cumulative File, 1972-2012 Coursera Extract.
Modified for Data Analysis and Statistical Inference course (Duke University).

R dataset could be downloaded at http://bit.ly/dasi_gss_data.

Original data:

Smith, Tom W., Michael Hout, and Peter V. Marsden. General Social Survey, 1972-2012
[Cumulative File]. ICPSR34802-v1. Storrs, CT: Roper Center for Public Opinion Research,
University of Connecticut /Ann Arbor, MI: Inter-university Consortium for Political
and Social Research [distributors], 2013-09-11. doi:10.3886/ICPSR34802.v1

Persistent URL: http://doi.org/10.3886/ICPSR34802.v1

### Other references

1. <a name="projecttemplate"/>ProjectTemplate, a system for automating the thoughtless
   parts of a data analysis project. URL: <http://projecttemplate.net/>.
2. <a name="gssfaq"/>General Social Survey (GSS) FAQ.
   URL: <http://publicdata.norc.org:41000/gssbeta/faqs.html>.
   Accessed 03/30/2014

Appendix
--------

Example of the data used in the study.

```{r study_data_head, warning=FALSE}
# First 50 observations
head(study_data, n=50L)
```