Exploration
===========

Project startup
---------------

Setting knitr options.

``` {r knitr_options}
require(knitr)
opts_knit$set(root.dir="..")
opts_chunk$set(fig.path="../graphs/expl-")
```

Load required libraries.
```{r load_libraries, warning=FALSE}
```
Load project with ProjectTemplate.

```{r load_project, warning=FALSE}
require(ProjectTemplate)
reload.project()
```

Exploring to choose proposal
----------------------------

Summary of all datatable, to explore candidate variables.
```{r gss_summary, warning=FALSE}
# summary(gss)
```

### First candidate analysis: Labor force status and Sex

Labor force status statistics.
```{r wrkstat, warning=FALSE}
wrkstat_t = table(gss$wrkstat, useNA="ifany")
wrkstat_t
par(mar = c(8, 4, 4, 2) + 0.1)
barplot(wrkstat_t, las=2, main="Labor force status")
```

Sex statistics.
```{r sex, warning=FALSE}
sex_t = table(gss$sex, useNA="ifany")
sex_t
par(mar = c(5, 4, 4, 2) + 0.1)
barplot(sex_t, las=2, main="Sex")
```

Labor force status and Sex.
```{r wrkstat_sex, warning=FALSE}
wrkstat_sex_t = table(gss$sex, gss$wrkstat, useNA="ifany")
wrkstat_sex_t
par(mar = c(5, 4, 4, 2) + 0.1)
mosaicplot(wrkstat_sex_t, las=2, main="Labor force status and Sex")
```

### Second candidate analysis: Highest degree and Income in constant dollar

Highest degree statistics.
```{r degree, warning=FALSE}
degree_t = table(gss$degree, useNA="ifany")
degree_t
par(mar = c(8, 4, 4, 2) + 0.1)
barplot(degree_t, las=2, main="Highest degree")
```

Family income in constant dollars.
```{r coninc, warning=FALSE}
summary(gss$coninc)
par(mar = c(5, 4, 4, 2) + 0.1)
hist(gss$coninc, main="Family income in constant dollars")
```

Family income in constant dollars by Degree.
```{r degree_coninc, warning=FALSE}
par(mar = c(8, 4, 4, 2) + 0.1)
boxplot(gss$coninc ~ gss$degree, las= 2, main="Family income in constant dollars by Degree")
```

Exploring for the proposal
--------------------------

### Create a dataset with the two variables chosen

Create a dataset with degree and coninc columns, then filter all the NAs.
Add descriptive columns names.
```{r proposal_ds, warning=FALSE}
proposal_ds = gss[,c(12,27)]
proposal_ds = proposal_ds[complete.cases(proposal_ds),]
colnames(proposal_ds) <- c("Highest.Degree","Family.Income.Constant.USD")
dim(proposal_ds)
head(proposal_ds, n=40L)
```

### Summary statistics for Highest Degree
Highest Degree is a categorical variable. Summarize with table and barplot.
```{r proposal_degree, warning=FALSE}
degree_table = table(proposal_ds$Highest.Degree)
degree_table
par(mar = c(8, 4, 4, 2) + 0.1)
barplot(degree_table, las=2, main="Highest Degree")
```

### Summary statistics for Family Income in constant USD
Family Income is a numerical continuous variable. Summarize with summary and histogram.
```{r proposal_conint, warning=FALSE}
summary(proposal_ds$Family.Income.Constant.USD)
par(mar = c(5, 4, 4, 2) + 0.1)
hist(proposal_ds$Family.Income.Constant.USD, main="Family Income in constant USD", xlab="USD")
```

### Are Family Income in constant USD dependent from Highest Degree?

```{r proposal_degree_coninc, warning=FALSE}
par(mar = c(8, 4, 4, 2) + 0.1)
boxplot(proposal_ds$Family.Income.Constant.USD ~ proposal_ds$Highest.Degree, las= 2, main="Family Income in constant USD by Degree")
```
